---
title: "Tidy Data"
output: github_document
---

```{r, echo = FALSE}
library(tidyverse)
library(readxl) #excel
library(haven) #sas
```
*gather and spread are older terms from tidyverse and not pivot longer (gather) pivot wider (spread)

```{r, eval = FALSE}

litters_df = read_csv("./data/FAS_litters.csv")
```


```{r}
litters_wide = 
  read_csv("./data/FAS_litters.csv") %>%
  janitor::clean_names() %>%
  select(litter_number, ends_with("weight")) %>% 
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd", 
    values_to = "weight") %>% 
  mutate(gd = recode(gd, "gd0_weight" = 0, "gd18_weight" = 18))
```


```{r}
analysis_result = tibble(
  group = c("treatment", "treatment", "placebo", "placebo"),
  time = c("pre", "post", "pre", "post"),
  mean = c(4, 8, 3.5, 4)
)
```

You see that below the time gets transposed to a time = pre --> pre becomes a column and time = post --> post becomes a column. the corresponding mean for each gets there


```{r}
pivot_wider(
  analysis_result, 
  names_from = "time", 
  values_from = "mean")
```
_Fellowship Data_


```{r}
fellowship_ring = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "B3:D6") %>%
  mutate(movie = "fellowship_ring")

two_towers = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "F3:H6") %>%
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "J3:L6") %>%
  mutate(movie = "return_king")
```

* using bind rows 

  * REQ: all of the columns in the datasets have the same names (race, female, movie)
  
```{r}
lotr_tidy = 
  bind_rows(fellowship_ring, two_towers, return_king) %>%
  janitor::clean_names() %>%
  pivot_longer(
    female:male,
    names_to = "gender", #formerly 1 col = female, 1 col = male --> 
    values_to = "words") %>% #formerly 1 value in a row denoting female and male frequency counts UNDER each respective column
  mutate(race = str_to_lower(race)) %>% 
  select(movie, everything()) 
```


### joining

```{r}
pup_data = 
  read_csv("./data/FAS_pups.csv") %>%
  janitor::clean_names() %>%
  mutate(
    sex = recode(sex, `1` = "male", `2` = "female"), #note that ` backquotepups_ tells R to output a number (vs "" which is for char)
    sex = factor(sex)) 
```

```{r}
litter_data = 
  read_csv("./data/FAS_litters.csv") %>%
  janitor::clean_names() %>%
  separate(group, into = c("dose", "day_of_tx"), sep = 3) %>%
  relocate(litter_number) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    dose = str_to_lower(dose))
```

## left join 

My intent is to analyze the baby 

* if there is a mom with a missing baby --> i don't need it 
* i want to join the mom (litter) (y) data TO the pup data (x)
* includes all rows of data x 
* (right join) includes all rows of data y

```{r}
fas_data = 
  left_join(pup_data, litter_data, by = "litter_number")
```



## anti join

* 
