---
title: "Homework 2"
author: "Jennifer Mizhquiri Barbecho"
output: github_document
---

```{r, echo = FALSE}
library(tidyverse)
library(readxl) #excel
library(haven) #sas
library(ggplot2)
```



# Problem 1

Read the data

```{r, echo = FALSE}

read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
```
```{r, message = FALSE}
nyctransit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>% # figure out how to make NAs blank
  select(line:entry, vending, ada) %>% 
  mutate(entry = recode(entry, "YES" = 1, "NO" = 0)) #when to use <-

```
```{r}
head(nyctransit_df)
```
This dataset contains administrative, compliance, and geographic data on the NYC transit system. It has `r ncol(nyctransit_df)` columns and `r nrow(nyctransit_df)` rows. At this point, the variable names have been cleaned and one variable (entry) has been recoded so that it is a 0 or 1 instead of a character. Other aspects of the data have not been further inspected so it is not tidy. 

_Additional information_

```{r addl_data, include = FALSE}
nyctransit_df %>%
  distinct(station_name, line) #is there a way for me to get the count in line


```
There are `r nrow(nyctransit_df %>% distinct(station_name,line))` distinct stations in the dataset. 
```{r}

nyctransit_df %>%
  filter(ada == TRUE) %>% 
  distinct(station_name, line) 
  
```

There are `r nrow(nyctransit_df %>% filter(ada == TRUE) %>% distinct(station_name, line))` stations that are ADA compliant. 

```{r}


  nyctransit_df %>%
  filter(vending == "NO") %>% 
  group_by(entry) %>% 
  summarize(vending_entrance = n())


#how to set it so that the false can be explicitly stated if I decided to mutate a variable 
#how to set it so that the code divides it for me as it didn't seem straightforward
```
 `r 69/(114+69)` % of stations with no vending allow entrance 

