---
title: "Homework 2"
author: "Jennifer Mizhquiri Barbecho"
output: github_document
---

```{r, echo = FALSE}
library(tidyverse)
library(readxl) #excel
library(haven) #sas
library(ggplot2)
```


Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).

Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?


Read the data

```{r, eval = FALSE}

read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
```
```{r}
nyctransit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>%
  select(line:entry, vending, ada) %>% 
  mutate(entry <- recode(entry, "YES" = 1, "NO" = 0))
```
```{r}
nyctransit_df
```

