---
title: "Homework 2"
author: "Jennifer Mizhquiri Barbecho"
output: github_document
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl) #excel

```



# Problem 1 (provided solution)


Below we import and clean data from `NYC_Transit_Subway_Entrance_And_Exit_Data.csv`. The process begins with data import, updates variable names, and selects the columns that will be used in later parts fo this problem. We update `entry` from `yes` / `no` to a logical variable. As part of data import, we specify that `Route` columns 8-11 should be character for consistency with 1-7.

```{r}
trans_ent = 
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>% 
  janitor::clean_names() %>% 
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

As it stands, these data are not "tidy": route number should be a variable, as should route. That is, to obtain a tidy dataset we would need to convert `route` variables from wide to long format. This will be useful when focusing on specific routes, but may not be necessary when considering questions that focus on station-level variables. 

The following code chunk selects station name and line, and then uses `distinct()` to obtain all unique combinations. As a result, the number of rows in this dataset is the number of unique stations.

```{r}
trans_ent %>% 
  select(station_name, line) %>% 
  distinct
```

The next code chunk is similar, but filters according to ADA compliance as an initial step. This produces a dataframe in which the number of rows is the number of ADA compliant stations. 

```{r}
trans_ent %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

To compute the proportion of station entrances / exits without vending allow entrance, we first exclude station entrances that do not allow vending. Then, we focus on the `entry` variable -- this logical, so taking the mean will produce the desired proportion (recall that R will coerce logical to numeric in cases like this).

```{r}
trans_ent %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```

Lastly, we write a code chunk to identify stations that serve the A train, and to assess how many of these are ADA compliant. As a first step, we tidy the data as alluded to previously; that is, we convert `route` from wide to long format. After this step, we can use tools from previous parts of the question (filtering to focus on the A train, and on ADA compliance; selecting and using `distinct` to obtain dataframes with the required stations in rows).

```{r}
trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct

trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

# Problem 2

_Step 1: Read and clean Mr. Trash Wheel dataset_
  * variable names are cleaned
  * rows not containing dumpster values are removed
  * note that "dumpster" is a numeric variable
  * sports_ball is rounded/set to an integer
  * note that year has been changed to an integer
  * a variable "wheel" with value of "mr" has been applied to all obs, to facilitate tracking after merging.

```{r mrwheels_chunk, echo = FALSE, message = FALSE}
mrwheels_df = 
  read_excel(
    "data/Trash Wheel Collection Data.xlsx", 
    sheet = "Mr. Trash Wheel", range = "A2:N549") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    wheel = "mr",
    sports_balls = as.integer(round(sports_balls)),
    year = as.integer(year)
  )
```


_Step 2: Professor Trashwheel dataset_ 
  * variable names are cleaned
  * rows not containing dumpster values are removed
  * a variable "wheel" with value of "prof" has been applied to all obs, to facilitate tracking and later merging
  * note that sports balls is NOT included as a variable
  * note that dumpster variable is a numeric variable
  * note that year is an integer


```{r profwheels_chunk, echo = FALSE, message = FALSE}
profwheels_df = 
  read_excel(
    "data/Trash Wheel Collection Data.xlsx", 
    sheet = "Professor Trash Wheel", range = "A2:M96") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>%
  mutate(wheel = "prof")
```


_Step 3: Combining Mr. Trash Wheel & Professor Trash Wheel Dataset_

  * Combine both datasets into a dataset called "combo_wheels_tidy"
  
```{r combo_trash_chunk, echo = FALSE, message = FALSE}
combo_wheels_tidy = 
  bind_rows(mrwheels_df, profwheels_df) %>%
  janitor::clean_names()
```

* About the Dataset:

```{r print_combo_trash}
head(combo_wheels_tidy)
```

There are `r nrow(combo_wheels_tidy)` observations in the resulting dataset. This dataset demonstrates various categories of collected items (i.e. sports balls, glass bottles, trash bottles) by weight (in tons), across years (`r combo_wheels_tidy %>% pull(year) %>% min()` - `r combo_wheels_tidy %>% pull(year) %>% max()`), for a trash-collection, water-vessel based initiative. This is helpful for generating summary stats on output and can help compare productivity between Mr. Trashwheels and Professor Trashwheels. For example, Mr. Trashwheel collected `r combo_wheels_tidy %>% filter(wheel == "prof") %>%  select(weight_tons) %>% sum()` tons, and in the year 2020, Mr. Trashwheels collected `r combo_wheels_tidy %>% filter( wheel == "mr", year == 2020) %>% select(sports_balls) %>% sum()` sports balls. 


# Problem 3

Import and clean pols-month.csv
 * note: data is sorted by year and month
 * year and month were moved to become the leading columns
 * month is a character variable and each value is renamed to their corresponding month
 * president variable was created to denote affiliation (i.e. dem = Democrat; gop = GOP)
 * FYI the dummy variable was dropped although the indicator variable of 0, 1 is false across select rows where they are listed as 2. This would require further inspection. 
 
```{r, echo = FALSE}
pols_df = 
  read_csv("data/fivethirtyeight_datasets/pols-month.csv") %>% 
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>% 
  mutate(
    year = as.integer(year),
    month = recode(month, "01" = "January", "02" = "February", "03" = "March", "04" = "April", "05" = "May", "06" = "June", "07" = "July", "08" = "August", "09" = "September", "10" = "October", "11" = "November", "12" = "December")) %>% 
  pivot_longer(
    cols = c('prez_dem','prez_gop'),
    names_to = "president", 
    names_prefix = "prez_",
    values_to = "dummy") %>% 
  select(-dummy, -day) %>% 
  select(year, month, everything()) %>% #does select always need to be broken down like this. yes.
  arrange(year, month)
  
```
```{r}
head(pols_df)
```

Import and clean snp.csv
 * note: data is sorted by year and month
 * year and month were moved to become the leading columns
 * month is a character variable and each value is renamed to their corresponding month
 * president variable was created to denote affiliation (i.e. dem = Democrat; gop = GOP)
 
```{r, echo = FALSE}
snp_df = 
  read_csv("data/fivethirtyeight_datasets/snp.csv") %>% 
  janitor::clean_names() %>%
  separate(date, into = c("month", "day", "year"), sep = "/") %>% 
  mutate(
    year = if_else(as.integer(year) < 23, 2000 + (as.integer(year)), 1900 + (as.integer(year))),
    month = recode(month, "1" = "January", "2" = "February", "3" = "March", "4" = "April", "5" = "May", "6" = "June", "7" = "July", "8" = "August", "9" = "September", "10" = "October", "11" = "November", "12" = "December")) %>%
  select(year, month, everything()) %>% 
  select(-day) %>% 
  arrange(year, month)

#you can try to use lubridate by using bracket and the dataset name 
#you can try to recode based on this by adding the 2000 to a certain range
```
```{r}
head(snp_df)
```

```{r, echo = FALSE}
unemp_df = 
  read_csv("data/fivethirtyeight_datasets/unemployment.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemp_freq") %>% 
  mutate(
    year = as.integer(year),
    month = recode(month, "jan" = "January", "feb" = "February", "mar" = "March", "apr" = "April", "may" = "May", "jun" = "June",  "jul" = "July", "aug" = "August", "sep" = "September", "oct" = "October", "nov" = "November", "dec" = "December" )) %>% 
  select(year, month, everything()) %>% 
  arrange(year, month)

```

```{r}
head(unemp_df)
```


_Joining SNP TO POLS_
* in which pols is the given x data set with `r ncol(pols_df)` columns
* in which snp is the given y data set with `r ncol(snp_df)` columns

```{r}
pols_snp_df = 
  left_join(pols_df, snp_df, by = c("year" = "year", "month" = "month"))
```
  
_Joining unemployment to the SNP/POLS merged dataset_

```{r}
pols_snp_unemp_df = 
  left_join(pols_snp_df, unemp_df, by =  c("year" = "year", "month" = "month"))
```

_Combined dataset_
```{r}
head(pols_snp_unemp_df)
```


The resulting combined dataset is `r ncol(pols_snp_unemp_df)` columns by `r nrow(pols_snp_unemp_df)` rows. The year span is `r pols_snp_unemp_df %>% pull(year) %>% min()` to `r pols_snp_unemp_df %>% pull(year) %>% max()`. Key data include the frequency of senators, governors, and representatives by party affiliation, the party affiliation of the president, the close number, and the unemployment rates, all by year and month.