---
title: "Homework 2"
author: "Jennifer Mizhquiri Barbecho"
output: github_document
---

```{r, echo = FALSE}
library(tidyverse)
library(readxl) #excel
library(ggplot2)
```



# Problem 1

Read the data

```{r, echo = FALSE}

read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
```
```{r, message = FALSE}
nyctransit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>% # figure out how to make NAs blank
  select(line:entry, vending, ada) %>% 
  mutate(entry = recode(entry, "YES" = 1, "NO" = 0)) #when to use <-

```
```{r}
head(nyctransit_df)
```
This dataset contains administrative, compliance, and geographic data on the NYC transit system. It has `r ncol(nyctransit_df)` columns and `r nrow(nyctransit_df)` rows. At this point, the variable names have been cleaned and one variable (entry) has been recoded so that it is a 0 or 1 instead of a character. Other aspects of the data have not been further inspected so it is not tidy. 

_Additional information_

```{r addl_data, include = FALSE}
nyctransit_df %>%
  distinct(station_name, line) #is there a way for me to get the count in line


```
There are `r nrow(nyctransit_df %>% distinct(station_name,line))` distinct stations in the dataset. 
```{r, echo = FALSE}

nyctransit_df %>%
  filter(ada == TRUE) %>% 
  distinct(station_name, line) 
  
```

There are `r nrow(nyctransit_df %>% filter(ada == TRUE) %>% distinct(station_name, line))` stations that are ADA compliant. 

```{r, echo = FALSE}


  nyctransit_df %>%
  filter(vending == "NO") %>% 
  group_by(entry) %>% 
  summarize(vending_entrance = n())


#how to set it so that the false can be explicitly stated if I decided to mutate a variable 
#how to set it so that the code divides it for me as it didn't seem straightforward
```
 `r 69/(114+69)` % of stations with no vending allow entrance.

_reformat the data_

```{r, eval = FALSE}
nyctransit_long = 
  nyctransit_df %>% 
  mutate(
    route8 = as.character(route8)) %>% 
  pivot_longer(
    nyctransit_df, route1:route11,
    names_to = "route_num",
    names_prefix = "route",
    values_to = "route_name") %>% 
  mutate(
    route_num = factor(route_num)
  )
  

```

# Problem 2

Mr. Trash wheel dataset. I am adding a variable wheel with value of mr applying to all obs

```{r}
mrwheels_df = 
  read_excel(
    "data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", 
    sheet = "Mr. Trash Wheel", range = "A2:N535") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    wheel = "mr",
    sports_balls = as.integer(round(sports_balls),
    dumpster = as.character(dumpster))
  )
```


Professor Trashwheel dataset. I am adding a variable wheel with value of prof and applying to all obs

* note that in professor wheels dumpster variable was numeric. to make it match what is in mr trashwheels, the variable is converted to character

_Office Hour question: How to make sure that I don't mistakenly add a variable to my original dataset_


```{r}
profwheels_df = 
  read_excel(
    "data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", 
    sheet = "Professor Trash Wheel", range = "A2:N117") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>%
  mutate(
    wheel = "prof",
    sports_balls = as.integer(round(sports_balls)),
    dumpster = as.character(dumpster)
  )
```

Combine both datasets into combo_wheels
```{r}
combo_wheels_tidy = 
  bind_rows(mrwheels_df, profwheels_df) %>%
  janitor::clean_names()
```

_About the combined dataset_

There are `r nrow(combo_wheels_tidy)` observations in the resulting dataset. This dataset demonstrates various categories of collected items (i.e. sports balls, glass bottles, trash bottles) by weight (in tons), across years (2014 - 2021). This is helpful for generating summary stats on output and can help compare efficacy between mr. trashwheels and professor trash wheels. For example, In the year 2020, Mr. Trashwheels collected `r combo_wheels_tidy %>% filter( wheel == "mr", year == 2020) %>% summarize(sports_balls = n())` sports balls. 

```{r}
combo_wheels_tidy %>% 
  filter(
    wheel == "mr",
    year == 2020) %>% 
  summarize(sports_balls = n())
```
```{r}
combo_wheels_tidy %>%
  distinct(year) #ask why this doesn't work in inline r
```

# Problem 3

Import and clean pols-month.csv
 * note: data is sorted by year and month
 * year and month were moved to become the leading columns
 * month is a character variable and each value is renamed to their corresponding month
 * president variable was created to denote affiliation (i.e. dem = Democrat; gop = GOP)
 * 
```{r}


pols_df = 
  read_csv("data/fivethirtyeight_datasets/pols-month.csv") %>% 
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>% 
  mutate(
    month = recode(month, "01" = "January", "02" = "February", "03" = "March", "04" = "April", "05" = "May", "06" = "June", "07" = "July", "08" = "August", "09" = "September", "10" = "October", "11" = "November", "12" = "December")) %>% 
  pivot_longer(
    cols = c('prez_dem','prez_gop'),
    names_to = "president", 
    names_prefix = "prez_",
    values_to = "dummy") %>% 
  select(-dummy, -day, year, month, everything()) %>% 
  arrange(year, month)
  
```

```{r}

```

